Composing like a human

### 2019-05-05 paper/main.tex
done with methods section

% Essentials
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[ruled]{algorithm2e}

@@

% Drawings
\usepackage{tikz}
\usetikzlibrary{arrows.meta, fit, positioning}

\tikzset{
    neuron/.style={
        circle, draw, thick,
        inner sep=0pt,
        minimum size=3.0em,
        node distance=4em
    },
    io/.style={
        neuron,
        fill=base01
    },
    conn/.style={
        thick,
        -{Straight Barb[angle=60:2pt 3]},
    }
}

@@

This chapter takes a deep-dive into the fields of meta-learning and generative networks in order to highlight their current state of affairs. This will lay the groundwork for the experiments detailed in Chapter \ref{chap:methods}. Section \ref{sec:lstm} will present the principal network architecture of this paper, the Long Short-Term Memory network, and argument for its use as the foundation of a generative network. Section \ref{sec:generative} will introduce Generative Adversarial Networks, a training method build specifically for the task, whereas Section \ref{sec:learn2learn} will formally introduce meta-learning, several approaches to it, and finally narrow down to current state-of-the-art algorithms.

\section{Long Short-Term Memory Networks} \label{sec:lstm}

In a standard ANN, the data traverses the network through the input layer, to the hidden layer, and finally reaches the output layer; setup which is generally known as a feedforward neural network \parencite[Figure \ref{fig:feedforward_net};][]{graves_supervised_2012}. Broadly, this type of network is defined by the fact that its connections do not form cycles. If however we relax this condition, we arrive at a recurrent neural network \parencite[Figure \ref{fig:rnn_folded};][]{rumelhart_learning_1986}: a family of neural networks specifically designed for processing sequential data \parencite{goodfellow_deep_2016}.

The information passes through the RNN in a similar manner to the feedforward network except that the activations which arrive at the hidden layer are from the current external input and prior hidden activations from the previous timestep \parencite{graves_supervised_2012}. Moreover, recurrent networks also benefit from the fact that parameters (weights) are shared across different parts of a model. This allows extending and applying the model to data points of different lengths and generalizing across all of them \parencite{goodfellow_deep_2016}. In contrast, a feedforward network would require separate parameters for each input feature, at each position. Thus, a formal definition of the network's forward pass would be
\begin{equation}
    h^t = f(h^{t-1}; x^t; \theta) \label{eq:rnn_hidden}
\end{equation}
where $h^t$ is the state of the hidden layer at time $t$, $x^t$ is the input at time $t$, and $\theta$ the bias.

The backward pass in a RNN is similar to that of a feedforward network: given the partial derivatives of a differentiable loss function $\mathcal{L}$ we calculate the derivatives with respect to the network's weights \parencite{graves_supervised_2012}. In this setting, a popular algorithm is called backpropagation through time

@@

@@

\begin{figure}[ht]
    \centering
    \begin{subfigure}{.24\linewidth}
        \begin{tikzpicture}
            [align=center]
            % Draw recurrent network
            \node[neuron] (h) {$h$};
            \node[io, below=4em of h] (x) {$x$};
            \node[io, above=4em of h] (y) {$y$};

@@

        \end{tikzpicture}
        \caption{Unfolded}
        \label{fig:rnn_unfolded}
    \end{subfigure}
\end{figure}

@@

Turning over to the generative side of things, nowadays, the most popular form of generating data is through the use of a Generative Adversarial Network \parencite[GAN;][]{goodfellow_generative_2014}. Here a generative model $G$ captures the data distribution and creates samples from it, whereas a discriminative model $D$ estimates the probability that the input is real (i.e., comes from the training data) rather than fake (i.e., comes from $G$). Formally, given $x$ sampled from real data, and $z$ being input noise, $D$ tries to keep $D(x)$ near $1$, while making $D(G(z))$ near $0$. Conversely, $G$ tries to make $D(G(z))$ near $1$ \parencite{goodfellow_nips_2016}. A notable implementation is C-RNN-GAN \parencite{mogren_c-rnn-gan_2016}, a continuous recurrent neural network with adversarial training. It consists of a network with two LSTM layers and 350 hidden units per layer, in both $G$ and $D$. However, $D$ has a bidirectional layout which allows is to take context from both past and future for its decisions, while $G$ is unidirectional \parencite{mogren_c-rnn-gan_2016}. As is the standard practice for GANs, $G$ will create fake samples of music, whereas $D$ will try to classify correctly which sample is real.

Notwithstanding GANs, other generative successes have made use of adaptation on the Long Short-Term Memory (LSTM) network \parencite{hochreiter_long_1997}. Here, the cells of a network are designed to allow for information to persist by implementing a recursive state unit, an input gate, a forget gate, and an output gate. The input of the network is allowed into the state unit if it is allowed to pass through by the input gate. Next, the forget gate controls the self-loop of the state cell. Finally, the output gate has the power to shut off the output cell \parencite{goodfellow_deep_2016}. Here, a notable implementation is Performance-RNN \parencite{oore_this_2018}; a three layer LSTM network with 512 layers each. Compared to C-RNN-GAN, here the network is completely unidirectional. Probably more important than the network architecture, a central piece of the model's performance is due to the final representation of the MIDI data: each time slice is represented as a one-hot encoded vector containing Note-On (128 values), Note-Off (128 values), Time-Shift (125 values), and Velocity (32 values) events \parencite{oore_this_2018}.

@@

\textcite{lake_human-level_2015} initially popularized the idea of learning from a few class examples using a probabilistic learning framework capable of generalizing and learning a large number of concepts from a single class example (i.e., one-shot learning). \textcite{rezende_one-shot_2016} were among the first to provide a solution to the one-shot problem in the context of a neural network (NN) by implementing Bayesian reasoning into a deep NN embedded within hierarchical latent variable models. The Reptile algorithm \parencite{nichol_first-order_2018} is a recent approach to this problem. It counteracts the shortcomings of gradient-optimization algorithms by learning the model's initial parameters in order to maximize its performance on novel tasks. Furthermore, it allows for any type of network to be used and does not place any restrictions on the type of loss function that can be used \parencite{nichol_first-order_2018}. The algorithm for Reptile is as follows:

\IncMargin{1em}
\begin{algorithm}
Initialize $\phi$, the vector of initial parameters

\For{iteration = $1, 2, \dots$}{
    Sample task $\tau$, corresponding to loss $L_{\tau}$ on weight vectors $\tilde \phi$

    Compute $\tilde \phi = U_{\tau}^{k} (\phi)$, denoting $k$ steps of SGD or Adam

    Update $\phi \leftarrow \phi + \epsilon (\tilde{\phi} - \phi)$
}

@@

Although simplistic, this algorithm reaches similar results to another popular few-shot algorithm called MAML \parencite{finn_model-agnostic_2017}. Both algorithms are initialization-based learning methods however, a major distinction between the two is that the vanilla MAML implementation requires calculating a second-order derivative, which can significantly slowdown computation \parencite{nichol_first-order_2018}. This forms a compelling reason to use Reptile as the few-shot learning method for this research.

@@

To provide an answer to the above research questions, this study will develop (in Keras) two generative models based on C-RNN-GAN and Performance-RNN, where the learning procedure will incorporate the Reptile algorithm. Other than that, the models will not differ from their original implementation. These models will firstly be compared to the baseline proposed by \textcite{larochelle_few-shot_2017} (a LSTM network trained on the entire train set) and then with the music from the dataset. The evaluation between the models with the baseline will be based on the number of statistically different bins \parencite[NDB;][]{richardson_gans_2018} and domain specific measurements (polyphony, scale consistency, repetitions, tone span). The evaluation between the model's generated samples and real music in the dataset will be based on the domain specific metric. Refer to the evaluation section for more details on these metrics.

\section{Dataset}

The dataset for this research will be similar to that of \parencite{oore_this_2018} namely, the MAESTRO Dataset \parencite{hawthorne_enabling_2018}. It consists of recorded MIDI data collected from each installment of the International Piano-e-Competition. All performances were done by piano experts on a Yamaha Disklavier, instrument which integrates a highly precise MIDI capture and playback system. Recorded MIDI events are of sufficient fidelity to allow judges to remotely listen to the contestant's performance (also on a Disklavier).

The dataset contains MIDI recordings from nine years of the International Piano-e-Competition. This amounts to 1,184 piano performances, approximately 430 compositions, 6.18 million notes played, and approximately 172 hours of playback. There is also a recommended train/validation/test split created on the following criteria \parencite{hawthorne_enabling_2018}:

@@

Moreover, each performance comes with additional metadata namely, the name of the composer, the title of the performance, the suggested train, validation, test splits, year of the performance, name of the file, and duration in seconds of the performance.

\section{Evaluation}

>>>>

% Essentials
\usepackage{tabularx}
\usepackage{supertabular}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[ruled]{algorithm2e}

@@

% Drawings
\usepackage{tikz}
\usetikzlibrary{arrows.meta, fit, positioning, shapes}
\newcommand{\empt}[2]{$#1^{\langle #2 \rangle}$}

\tikzset{
    neuron/.style={
        circle, draw, thick,
        inner sep=0pt,
        minimum size=3.0em,
        node distance=4em
    },
    io/.style={
        neuron,
        fill=base01
    },
    conn/.style={
        thick,
        -{Straight Barb[angle=60:2pt 3]},
    },
    connl/.style={
        thick,
        {Straight Barb[angle=60:2pt 3]}-,
    },
    prod/.style={circle, draw, thick, inner sep=0pt},
    ct/.style={circle, draw, inner sep=5pt, ultra thick,
		minimum width=10mm},
    ft/.style={circle, draw, thick, minimum width=8mm, inner sep=1pt},
    filter/.style={circle, draw, thick, minimum width=7mm, inner sep=1pt,
		path picture={
			\draw[thick, rounded corners] (path picture bounding box.center)--++(65:2mm)--++(0:1mm);
			\draw[thick, rounded corners] (path picture bounding box.center)--++(245:2mm)--++(180:1mm);}},
}

@@

This chapter takes a deep-dive into the fields of meta-learning and generative networks in order to highlight their current state of affairs. This will lay the groundwork for the experiments detailed in Chapter \ref{chap:methods}. Section \ref{sec:rnn} will present the principal network architecture of this paper, the recurrent neural network, and argument for its use as the foundation of a generative network. Section \ref{sec:generative} will introduce Generative Adversarial Networks, a training method build specifically for the task, whereas Section \ref{sec:learn2learn} will formally introduce meta-learning, several approaches to it, and finally narrow down to current state-of-the-art algorithms.

\section{Recurrent Neural Networks} \label{sec:rnn}

In a standard ANN, the data traverses the network through the input layer, to the hidden layer, and finally reaches the output layer; setup which is generally known as a feedforward neural network \parencite[Figure \ref{fig:feedforward_net};][]{graves_supervised_2012}. Broadly, this type of network is defined by the fact that its connections do not form cycles. If however we relax this condition, we arrive at a recurrent neural network \parencite[Figure \ref{fig:rnn_folded};][]{rumelhart_learning_1986}: a family of neural networks specifically designed for processing sequential data \parencite{goodfellow_deep_2016}. It is possible to use a feedforward architecture, specifically a one-dimensional Convolutional Neural Network, with sequential data, where the convolution would be the basis for building a time-delay neural network. Unfortunately, the output of such a network would be a function of a small number of neighbouring data points \parencite[say, the last three notes;][]{goodfellow_deep_2016} whereas in the contex of music, knowledge of the notes played from the begining of the song is needed. As such, RNNs seem to be a more compelling alternative for this task.

The information passes through the RNN in a similar manner to the feedforward network except that the activations which arrive at the hidden layer are from the current external input and prior hidden activations from the previous timestep \parencite{graves_supervised_2012}. Moreover, recurrent networks also benefit from the fact that parameters (weights) are shared across different parts of a model. This allows extending and applying the model to data points of different lengths and generalizing across all of them \parencite{goodfellow_deep_2016}. In contrast, a feedforward network would require separate parameters for each input feature, at each position. Thus, a formal definition of the network's forward pass would be
\begin{equation}
    h^t = f(h^{t-1}; x^t; \theta) \label{eq:rnn_hidden}
\end{equation}
where $h^t$ is the state of the hidden layer at time $t$, $x^t$ is the input at time $t$, and $\theta$ the model parameters.

The backward pass in a RNN is similar to that of a feedforward network: given the partial derivatives of a differentiable loss function $\mathcal{L}$ we calculate the derivatives with respect to the network's weights \parencite{graves_supervised_2012}. In this setting, a popular algorithm is called backpropagation through time \parencite[BPTT;][]{werbos_backpropagation_1990} which consists of repeated applications of the chain rule. The difference between standard backpropagation and BPTT is that the loss function depends on the activation of the hidden layer through its influence on the output layer as well as its influence on the hidden layer a the next timestep \parencite{graves_supervised_2012}.

@@

The reason why these notions have been introduced is to be able to bring up the following point. As good as standard RNNs are - in theory - at accessing contextual information when mapping input sequences to output sequences, the range of contexts that can be accessed in a standard RNN architecture is in practice quite limited \parencite{graves_supervised_2012}. \textcite{hochreiter_long_1997} outlines two problems caused by BPTT: the error signals flowing bacwards in time tend to either (1) blow up or (2) vanish. As such, the Long Short-Term Memory network has been proposed, a recurrent architecture designed to overcome these issues \parencite{hochreiter_long_1997}.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}{.24\linewidth}
        \begin{tikzpicture}
            [align=center]
            % Draw recurrent network
            \node[neuron] (h) {$h$};
            \node[io, below=4em of h] (x) {$x$};
            \node[io, above=4em of h] (y) {$y$};

@@

        \end{tikzpicture}
        \caption{Unfolded}
        \label{fig:rnn_unfolded}
    \end{subfigure}
    \caption{A typical view of an RNN (left) and an unfolded view of the same network (right). In both sub-figures each node is a layer of network units however, in the second representation these are shown for each timestep. The weighted connections from the input layer to the hidden one is labelled $U$, those from the hidden layer feeding back to itself are $W$, and those from the hidden layer to the output layer are labelled as $V$. Note that these weights are reused at each timestep. Bias has been omitted in this representation. Figure adapted from \textcite{graves_supervised_2012}}
\end{figure}

The LSTM is made of a network of recurrently connected subnetworks known as memory blocks, each with one or more self-connected memory cells and three multiplicative units: the input gate, the output gate, and the forget gate \parencite{hochreiter_long_1997}. The three gates act as analogies for the read, write, reset computer operations and allow the memory cells to store and access information over long periods of time \parencite{graves_supervised_2012}. See Figure \ref{fig:lstm} for an illustration of a LSTM memory block with a single cell.

Aside from being successful at a variety of classification tasks \parencite[e.g.,][]{graves_framewise_2005}, this architecture has been successfully applied to generating new data points \parencite[such as creating sequences of text;][]{graves_generating_2013}. In the context of generating musical performances, experiments with this neural architecture go as far as \textcite{eck_first_2002} however, a more noteworthy implementation is Performance-RNN \parencite{oore_this_2018}; a three layer LSTM network with 512 cells each designed to generate classical compositions whilst maintaining the playing style of expert pianists. Considering how successful it was, this model will be re-implemented in this paper's experiments (see Chapter \ref{chap:methods}).

\begin{figure}[t]
    \centering
    \begin{tikzpicture}
        \node[ct, label={Cell}] (ct) {$c^t$};
        \node[filter, right=of ct] (int1) {};
        \node[prod, right=of int1] (x1) {$\times$}; 
        \node[right=of x1] (ht) {$h^t$};
        \node[prod, left=of ct] (x2) {$\times$}; 
        \node[filter, left=of x2] (int2) {};
        \node[prod, below=5mm of ct] (x3) {$\times$}; 
        \node[ft, below=5mm of x3, label={right:Forget Gate}] (ft) {$f^t$};
        \node[ft, above=of x2, label={right:Input Gate}] (it) {$i^t$};
        \node[ft, above=of x1, label={right:Output Gate}] (ot) {$o^t$};

        \foreach \i/\j in {int2/x2, x2/ct, ct/int1, int1/x1,
                        x1/ht, it/x2, ct/it, ct/ot, ot/x1, ft/x3}
                            \draw[->, conn] (\i)--(\j);

                            \draw[->, conn] (ct) to[bend right=45] (ft);

                            \draw[->, conn] (ct) to[bend right=30] (x3);
                            \draw[->, conn] (x3) to[bend right=30] (ct);

                            \node[fit=(int2) (it) (ot) (ft), draw, inner sep=0pt] (fit) {};

                            \draw[<-, connl] (fit.west|-int2) coordinate (aux)--++(180:7mm) node[left]{$x^t$};
                            \draw[<-, connl] ([yshift=1mm]aux)--++(135:7mm);
                            \draw[<-, connl] ([yshift=-1mm]aux)--++(-135:7mm);

                            \draw[<-, connl] (fit.north-|it) coordinate (aux)--++(90:7mm) node[above]{$x^t$};
                            \draw[<-, connl] ([xshift=1mm]aux)--++(45:7mm);
                            \draw[<-, connl] ([xshift=-1mm]aux)--++(135:7mm);

                            \draw[<-, connl] (fit.north-|ot) coordinate (aux)--++(90:7mm) node[above]{$x^t$};
                            \draw[<-, connl] ([xshift=1mm]aux)--++(45:7mm);
                            \draw[<-, connl] ([xshift=-1mm]aux)--++(135:7mm);

                            \draw[<-, connl] (fit.south-|ft) coordinate (aux)--++(-90:7mm) node[below]{$x^t$};
                            \draw[<-, connl] ([xshift=1mm]aux)--++(-45:7mm);
                            \draw[<-, connl] ([xshift=-1mm]aux)--++(-135:7mm);
    \end{tikzpicture}
    \caption{LSTM memory block with one cell. Here the gates are nonlinear summation units that collect activations from the inside and outside of the memory block and control the activation of the cell $c^t$ via multiplications. The input and output gates ($i^t$, $o^t$) multiply the block's input and output activations respectively, while the forget gate ($f^t$) multiplies the previous hidden state. The input and output activation are typically tanh or sigmoid. The only outputs ($h^t$) from the block come from the output gate multiplication. Figure adapted from \textcite{graves_supervised_2012}}
    \label{fig:lstm}
\end{figure}

@@

Nowadays, the most popular form of generating data is through the use of a Generative Adversarial Network \parencite[GAN;][]{goodfellow_generative_2014}. The training procedure is set up as a game between two networks: a generator and a discriminator. The purpose of the former is to create samples that come from the same distribution as the training data, whereas the latter aims to determine which samples are real or fake \parencite{goodfellow_nips_2016}.

To make the generator's data distribution $p_g$ imitate that of the real data $p_{\text{data}}$, we need to define a prior on input noise variables $p_{z}(z)$ and then represent a mapping to the data space $G(z; \theta_g)$ where $G$ is a differentiable function represented by a network with parameters $\theta_g$. The discriminator is another network with a mapping function $D(x; \theta_d)$ that outputs a single scalar where $D(x)$ is the probability that $x$ came from the real data distribution rather than that of the generator \parencite{goodfellow_nips_2016}. We train $D$ to maximize the probability of correctly assigning a label to the training data and samples from $G$ ("real" and "fake"), while simultaneously training $G$ to minimize the dissimilarity between the two data distributions, without actually looking at $p_{\text{data}}$. In other words, $G$ must minimize $\log (1- D(G(z)))$ \parencite{goodfellow_generative_2014}.

In other words, both models play a minimax game with a value function $V(G, D)$:
\begin{equation}
    \min_G \max_D V(D, G) = \mathbb{E}_{x ~ p_{\text{data}}} [ \log D(x) ] + \mathbb{E}_{z ~ p_z (z)} [ \log (1-D(G(z))) ]
\end{equation}
where the solution is for $G$ to recover the training data distribution and $D$ to have a classification accuracy of 50\% \parencite{goodfellow_generative_2014}.

Although new, GANs are a significant conceptual and technical innovation \parencite{briot_deep_2017} which can be seen in them revitalizing interest in generative models. As mentioned before, they have been proven successful at image super-resolution \parencite{ledig_photo-realistic_2016}, image-to-image translation \parencite{isola_image--image_2016}, or style-based image generation \parencite{karras_style-based_2018}. Having said these, \textcite{mogren_c-rnn-gan_2016} has shown that GANs do not belong only to the field of Computer Vision by re-purposing this architecture to generate music. The model in question, C-RNN-GAN, features a completely unidirectional LSTM network with two layers, 350 units each as generator and a unidirectional LSTM network with a similar structure as discriminator.

@@

The rationale for investing time and energy into meta-learning is that we cannot capture the essence of learning simply by relying on a small number of algorithms, but rather a bunch of context-dependent learning strategies that would enable us to properly capture domain specific information \parencite{schmidhuber_evolutionary_1987}. Given the complexities of each learning strategy, \textcite{schmidhuber_evolutionary_1987} argues that an obvious way to tackle this initiative is to give a system the ability to learn how to learn. A simplistic way to do this would be to train a system on different tasks (e.g, image classification, text classification, etc.). However, this approach entails training on a dataset that is much larger that what current hardware can withstand. Furthermore, this approach would stray away from the idea of building general-purpose A.I. architectures as generalizing over a task would require more data points that what a human would need. To be more concrete, a human is able to generalize a task/object after encountering one (or few) example(s) \parencite{ravi_optimization_2016, chen_closer_2018}. Therefore, besides having to perform on a variety of tasks, a systems would need to excel at that using sparse amounts of data.

Although ideas of meta-learning can go as far as Schmidhuber's thesis in 1987, it seems that training a network using a handful of examples did not get much traction until \textcite{lake_human-level_2015} successfully designed a probabilistic learning framework capable of generalizing and learning a large number of concepts from a single class example (i.e., one-shot learning). For clarity, such a problem is typically defined as a $n$-shot $k$-way problem, where $n$ stands for the number of examples per class, and $k$ is the number of classes to be used in the training process.

Initially, it was thought that such a framework would not work in the context of an ANN since the iterative nature of gradient-based optimization algorithms do work well in the context of a set number of updates \parencite{ravi_optimization_2016} - and even more so when there are only a few examples to work with. Having said that, \textcite{rezende_one-shot_2016} were among the first to provide a solution to the one-shot problem in the context of a neural network by implementing Bayesian reasoning into a deep network embedded within hierarchical latent variable models. Since that, numerous efforts were made to develop an algorithm that enables an ANN to perform with a constrained number of class examples. Probably the most popular algorithm at the time of writing is the Model-Agnostic Meta-Learning (MAML) algorithm \parencite{finn_model-agnostic_2017}. One of MAML's core mechanics is the parameter update by calculating second-order derivatives. This procedure however implies a substantial computational penalty, which the authors did note. Therefore, they developed a first-order variant of MAML (FOMAML) which is able to perform similarly, if only slightly worse \parencite{finn_model-agnostic_2017}.

This however, seemed to be the initial thought, until \textcite{nichol_first-order_2018} showed that FOMAML tends to be somewhat unstable and very sensitive to even the smallest changes in hyper-parameters. Therefore, they developed Reptile, a first-order meta-learning algorithm that rivals both MAML and FOMAML \parencite{nichol_first-order_2018}. Just like MAML, it counteracts the shortcomings of gradient-optimization algorithms by learning the model's initial parameters in order to maximize performance on novel tasks, it allows for any type of network to be used, and does not place any restrictions on the type of loss function that can be used \parencite{nichol_first-order_2018}. The difference between the two however, lies in the way the parameters are updated.

\IncMargin{1em}
\begin{algorithm}
Initialize $\theta$, the vector of initial parameters

\For{iteration = $1, 2, \dots$}{
    Sample task $\tau$, corresponding to loss $L_{\tau}$ on weight vectors $\tilde \theta$

    Compute $\tilde \theta = U_{\tau}^{k} (\theta)$, denoting $k$ steps of SGD or Adam

    Update $\theta \leftarrow \theta + \epsilon (\tilde{\theta} - \theta)$
}

@@

In essence, the way Reptile works (see Algorithm \ref{alg:reptile}) is as follows. A model with parameters $\theta$ is initialized, and a meta-loop (or meta-epochs) of $i$ iterations starts. Within it, a task $\tau$ (e.g, image classification, text classification, etc.) is sampled, along with its corresponding loss function $L_{\tau}$ (for clarity we assume that each type of task uses a different loss function). Then we compute the updated parameters $\tilde \theta$ in an inner-loop with $k$ iterations on task $\tau$ (or in other words, once we sampled our task $\tau$, we train our model $k$ times on it using the $\theta$ as parameters so we can get $\tilde \theta$). Finally, we update out initial parameters $\theta$ using this formula: $\theta \leftarrow \theta + \epsilon (\tilde{\theta} - \theta)$. Then the algorithm continues to the next iteration of the meta-loop carrying over our updated parameters $\theta$.

The explanation above is fairly verbose however, the algorithm itself is fairly simplistic but achieves similar performances to FOMAML \parencite{nichol_first-order_2018}. This forms a compelling reason to use Reptile as the few-shot learning method of choice for this paper.

@@

To reiterate from Section \ref{sec:thesis_goal}, the purpose of this paper is to establish (1) the extent to which the music generated under a few-shot learning algorithm is comparable to that of a model trained on the entire dataset, and (2) the extent to which the music from such a model is comparable to that of real musical performances. For these purposes, Performance-RNN and C-RNN-GAN were adapted to the Reptile training procedure (Algorithm \ref{alg:reptile}) and evaluated against the performances generated by the baseline proposed in \textcite{larochelle_few-shot_2017} - a single layer LSTM with 256 units - and actual musical performances from the dataset (Section \ref{sec:dataset}). For an in-depth explanation on the steps required to develop a model that is able to generate new data points refer to Section \ref{sec:how_to_gen} and to Section \ref{sec:reptile_train} for details on the training procedure. The evaluation of the models' performances will be based on the number of statistically different bins \parencite[NDB;][]{richardson_gans_2018} and several domain specific measurements adapted from \textcite{mogren_c-rnn-gan_2016}: polyphony, scale consistency, repetitions, tone span. Refer to Section \ref{sec:evaluation} for more details on these metrics.

\section{Dataset} \label{sec:dataset}

The dataset for this research will be similar to that of \parencite{oore_this_2018} namely, the MAESTRO Dataset \parencite{hawthorne_enabling_2018}. It consists of recorded MIDI data collected from each installment of the International Piano-e-Competition. All performances were done by piano experts on a Yamaha Disklavier, instrument which integrates a highly precise MIDI capture and playback system. The claim is that the recorded MIDI events are of sufficient fidelity to allow judges to remotely listen to the contestant's performance (also on a Disklavier).

The dataset contains MIDI recordings from nine years of the International Piano-e-Competition. This amounts to 1,184 piano performances, approximately 430 compositions, 6.18 million notes played, and approximately 172 hours of playback. There is also a recommended train/validation/test split (954, 105, 125 songs respectively) created on the following criteria \parencite{hawthorne_enabling_2018}:

@@

Moreover, each performance comes with additional metadata such as, the name of the composer, the title of the performance, the suggested train, validation, test splits, year of the performance, name of the file, and duration in seconds of the performance. Having said that, the dataset lacks clearly defined classes. The canonical composer could be used as a substitute however, the author strongly believes that there is a better way to categorize music: by genre.

Broadly speaking, Western classical music can be segregated into a number of genres based on the year of the composition and style of the piece. For this we can use the Concise Oxford Dictionary of Music \parencite{kennedy_concise_2007} which indicates to the following classical genres: baroque (1600 - 1750), classical (1750 - 1820), romanticism (1780 - 1910), modernism (1890 - 1930), and impressionism (circa 1890 - 1925). Granted, there are far more genres, but these are the ones that are covered by the dataset. As such, given the definitions of \textcite{kennedy_concise_2007} on said genres and artists, each individual piece from our dataset was manually assigned a genre. Concretely, the style of music was firstly determined by the period in which the composer has lived (in baroque, classical, etc.) or - if the genres would overlap (i.e., modernism, and impressionism) - the composer's primary style of music would be chosen according to the definition of the composer given by \textcite{kennedy_concise_2007}. Said labelling can be found in Appendix \ref{appendix:labelling}.

\section{Data Processing} \label{sec:data_proc}

\begin{figure}[t]
    \centering
    \begin{verbatim}
<note_on note=69 velocity=73 time=33>
<note_on note=71 velocity=78 time=15>
<control_change control=64 value=83 time=27>
<note_off note=69 velocity=0 time=6>
<note_off note=71 velocity=0 time=3>
    \end{verbatim}
    \caption{Example of MIDI messages in the dataset, as printed by the Python library Mido \parencite{bjorndalen_mido_2018}. At 33 ticks since the last event, the player pressed an A4 (note 69), followed by a B4 (note 71), 15 ticks later. Next, the pedal (Control Change message with a value of 64) is pressed 27 ticks later, and the keys are released 6 and 3 tick later.}
    \label{fig:midi_examples}
\end{figure}

The dataset consists of MIDI files which would require some processing in order to be ready for a LSTM network. Here, the approach of \textcite{oore_this_2018} was adopted where MIDI midi messages are transformed into a sequence of one-hot encoded events. More specifically, MIDI excerpts are represented as a sequence of events from a vocabulary of 416 different events - similar to \textcite{oore_this_2018}, but with three extra events in the vocabulary:
\begin{itemize}
    \item 128 Note-On events: starts a new note (one for each of the 128 MIDI pitches)
    \item 128 Note-Off events: releases a note (one for each of the 128 MIDI pitches)
    \item 126 Time-Shift events: moves the time step forward somewhere between 0 ms and 1 second, in increments of 8 ms
    \item 32 Velocity events: changes the velocity for each subsequent note, until the next velocity event
    \item 2 Pedal events: triggers or releases the pedal
\end{itemize}

A glance over the MIDI documentation \parencite{midi_association_official_2019} shows how complex and granular the standard is. However, we do not need to implement all of the messages into our model in order to make it work. We only need three: \texttt{note\_on}, \texttt{note\_off}, \texttt{pedal\_on/off}. Figure \ref{fig:midi_examples} show an example of how the midi messages look like. It should be noted that in this figure, the time is represented in ticks, MIDI's smallest unit of time, who's value is usually a third of a beat \parencite{midi_association_official_2019}.

As it can be seen in Figure \ref{fig:midi_examples}, each message contains multiple pieces of information such as the message type, the note/value of the message, the delta time (time elapsed since last message and the current message, measured in ticks), and velocity (if applicable). To one-hot encode this stream of messages, it is necessary to separate it so it contains only one type of information per entry. Thus, the sequence can be further broken down such that the first message that the system encounters is the delta time, followed by the velocity, and the note/pedal event (and their respective values). An example of how the MIDI sequences are broken down into singular pieces of information can be found in Figure \ref{fig:representation_midi}.

\begin{figure}[t]
    \begin{tikzpicture}
        \draw[step=1cm,base01,very thin] (0,0) grid (6,6);
        \filldraw[fill=base01, draw=base01] (0,1) rectangle (1,2);
        \filldraw[fill=base01, draw=base01] (2, 3) rectangle (5, 4);
        \draw (6.5, 5) node[anchor=south west, color=base09] {1. Move time by 1 tick};
        \draw (6.5, 4) node[anchor=south west, color=base0A] {2. Set velocity to 80};
        \draw (6.5, 3) node[anchor=south west, color=base0B] {3. Note 3 on};
        \draw (6.5, 2) node[anchor=south west, color=base0C] {4. Move time by 3 ticks};
        \draw (6.5, 1) node[anchor=south west, color=base0D] {5. Note 3 off};

        \draw[ -{Straight Barb[angle=60:2pt 3]}, color=base09, thick] (1, 1) -- (2, 1) node[anchor=north, midway] {1};
        \draw[color=base0A, thick] (2, 1) -- (2, 3) node[anchor=east, midway] {2};
        \draw[color=base0B, thick] (2, 3) -> (2, 4) node[anchor=east, midway] {3};
        \draw[ -{Straight Barb[angle=60:2pt 3]}, color=base0C, thick] (2, 3) -- (5, 3) node[anchor=north, midway] {4};
        \draw[color=base0D, thick] (5, 3) -- (5, 4) node[anchor=west, midway] {5};

        \draw[thick,-{Straight Barb[angle=60:2pt 3]}] (0,0) -- (6.5,0) node[anchor=north west] {Time (in ticks)};
        \draw[thick,-{Straight Barb[angle=60:2pt 3]}] (0,0) -- (0,6.5) node[anchor=south east] {Notes};
        \foreach \x in {0,1,2,3,4,5,6}
           \draw (\x cm,1pt) -- (\x cm,-1pt) node[anchor=north] {$\x$};
        \foreach \y in {0,1,2,3,4,5}
            \draw (1pt,\y.5 cm) -- (-1pt,\y.5 cm) node[anchor=east] {$\y$};
    \end{tikzpicture}
    \caption{Example of how the data is represented. The progression illustrates how a piano roll of MIDI data is converted into a sequence of commands (right hand side) that belong to our events vocabulary. Figure adapted from \textcite{oore_this_2018}.} 
    \label{fig:representation_midi}
\end{figure}

Having said these, it should be mentioned that the one-hot representation of MIDI messages has undergone some fine quantization, again, in a similar fashion to \textcite{oore_this_2018}. The delta times have been modified to correspond to a change of multiples of 8 ms (i.e., 0 ms, 8 ms, 16 ms, etc.) and the velocities have been partitioned into 32 steps as opposed to MIDI's 128. With regards to the time movements, \textcite{friberg_overview_2006} note that a noticeable difference in temporally displacing a single tone in a sequence takes place from 10 ms and above. When in comes to the note velocities, it should be pointed to the reader that classical music features eight levels of common dynamic marking (from $ppp$ to $fff$) which is why \textcite{oore_this_2018} note that 32 steps are more than enough.

\section{You're a musician Schmidhuber} \label{sec:how_to_gen}

Knowing what we know about RNNs and LSTMs from Section \ref{sec:rnn} one can devise a method where the network can generate a sequence of events. All that is needed is to present the system with an event to kickstart the generative process. Namely, an input $x^t$ of size 1 is given to the network to output the next likely event $x^{t+1}$, which in turn is fed back to generate $x^{t+2}$. This process is repeated until we have a sequence of events of a desired length.

An alternative to this would be to define a noise vector $z^t$ of an arbitrary size (e.g., 100), and a random input $x^t$ selected from our events vocabulary. The noise vector is passed through a linear layer that output a tensor that would serve as the network's hidden state $h^{t}$ at time $t$. It is then that $x^t$ and $h^{t}$ are fed into the LSTM layer to generate $x^{t+1}$ and $h^{t+1}$. The reason for pre-defining the hidden state for the first input is to increase the number of possible outcomes coming from the network.

To be more precise, if we were to present the network with only $x^t$ then the network would mark the input sequence as the beginning of the song. With that information, it is possible that the network would affix itself to following a specific pattern. For instance, if the first input would be a C\#, the network could learn to always output E as the next note. Using a random hidden state is a measure that was included to ensure some diversity in the process of generating samples (granted, it should be admitted the possibility of this scenario being unlikely, and that the regular method described above would still work just as good).

Having said this, diversity is not the only reason for including a noise vector $z$ into the training process. The reader might recall that this is part of the generation process in the framework of a GAN respectively, how the generator $G$ would create samples. Hence, all three models implemented in this paper feature the architecture of a GAN generator.

In terms of generating the next event, \textcite{oore_this_2018} have defined a parameter $\tau$ called temperature \parencite{google_magenta_performance_2017}. It is a value which uses the model's predicted events distribution to control the randomness of the output samples. In other words, let $\phi$ be the output of a LSTM network (for clarity, we ignore the hidden state $h$). Then, the next sequence $x^{t+1}$ is defined as 
\begin{equation}
    x^{t+1} = X(\omega)
\end{equation}
where
\begin{equation}
    X = \text{softmax}(\phi / \tau).
\end{equation}
$X$ is the predicted event distribution of the network's output divided by our temperature $\tau$. Thus, $x^{t+1}$ is actually a random sample from our defined categorical distribution $X$.

\section{How to train a Reptile} \label{sec:reptile_train}
